---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# TrustVDJ

The goal of TrustVDJ is to read and analyze immune repertoire data, especially from TRUST4, 10x Genomics cellranger or AIRR format results.

## Installation

1. TrustVDJ is available on CRAN:

``` r
install.packages('TrustVDJ')
```

2. Install TrustVDJ using devtools:

``` r
devtools::install_github('HatsuneCode/TrustVDJ')
```

*** Maybe dependency 'Biostrings' is not available:
``` r
install.packages('BiocManager')
BiocManager::install('Biostrings')
```

## Version

- v0.1.0: initial release.
- v0.1.1: fix bug in build_IMGT_reference: when same gene name, combine their sequences.

## Example

There are some basic examples showing how to read 10x/TRUST4 data commonly:

```{r example}
library(TrustVDJ)
## basic example code

# 10x cellranger:
airr10x   = system.file('extdata', '10x_airr_rearrangement.tsv.gz', package = 'TrustVDJ')
contig10x = system.file('extdata', '10x_filtered_contig_annotations.csv.gz', package = 'TrustVDJ')
vdj10x    = Read10x(airr_file = airr10x, contig_file = contig10x, verbose = FALSE)
summary(vdj10x[,1:3])

# TRUST4:
airrTrust = system.file('extdata', 'TRUST4_airr.tsv.gz', package = 'TrustVDJ')
bcTrust   = system.file('extdata', 'TRUST4_barcode_report.tsv.gz', package = 'TrustVDJ')
vdjTrust  = ReadTrust(airr_file = airrTrust, barcode_report_file = bcTrust, verbose = FALSE)
summary(vdjTrust[,1:3])
```

## Demo analyze
0. Library
```{r}
# detach('package:TrustVDJ', unload = T)
# devtools::install_github('HatsuneCode/TrustVDJ')
suppressMessages(library(TrustVDJ))
```
1. Perpare data
```{r}
# setwd('D:/TrustVDJ')
name    = c('GSM5160432_H1 GSM5160434_H2 GSM5160435_H3', 
            'GSM5160417_P1 GSM5160420_P2 GSM5160422_P3 GSM5160424_P4 GSM5160427_P5 GSM5160430_P6',
            'CMV COVID-19 EBV HSV-2 InfluenzaA')
name    = unlist(strsplit(name, ' '))
samples = sub('.*_', '', name)
groups  = list(H = samples[1:3], P = samples[4:9], V = samples[10:14])
paths   = system.file('extdata', 
                      c(paste0(name[1:9], '_TCR_filtered_contig_annotations.csv.gz'),
                        paste0(name[10:14], '.virus.tsv.gz')), 
                      package = 'TrustVDJ')
rm(name)
print(groups)
```
2. Read data
```{r}
property = list( ConsenID = c('raw_consensus_id'),
                 ClonoID  = c('raw_clonotype_id', 'complex.id'),
                 Vgene    = c('v_gene', 'V'),
                 Jgene    = c('j_gene', 'J'),
                 CDR3aa   = c('cdr3', 'CDR3'),
                 Barcode  = c('barcode', 'complex.id') )
samples  = lapply(seq(paths), function(i) {
  cat('--> read:', samples[i], '<--\n')
  data = Read10x(contig_file = paths[i], verbose = F)
  if('complex.id' %in% colnames(data))
    data$complex.id[ data$complex.id == 0 ] = make.unique(as.character(
      data$complex.id[ data$complex.id == 0 ] ))
  CreateVdjSample(data, property, name = samples[i], rm_allele = T, verbose = F)
})
vdj = CreateVdjObject(samples, groups, verbose = F)
rm(samples, property)
# saveRDS(vdj, 'vdj.object.rds')
print(vdj)
```
3. Filter vdj
Table VDJ genes:
```{r}
P_gene = TableVDJ(vdj, names = vdj@info$P, out.pref = 'P')
H_gene = TableVDJ(vdj, names = vdj@info$H, out.pref = 'H')
head(H_gene)
```
Filter > 2 sample:
```{r}
P_gene = P_gene$Gene[P_gene$TotalName > 2]
H_gene = H_gene$Gene[H_gene$TotalName > 2]
# venn
venn   = VennDiagram::venn.diagram(list(P = P_gene, H = H_gene),
                                   col             = scales::hue_pal()(2),
                                   margin          = 0.05,
                                   fontfamily      = 'sans',
                                   cat.fontfamily  = 'sans',
                                   filename        = NULL,
                                   disable.logging = T)
grid::grid.draw(venn)
dev.off()
# pie
pie = Pie(P_gene, H_gene) + Pie(H_gene, P_gene, fill = c('white', '#00BFC4'))
# ggsave('P-H.VDJ.uniquePie.pdf', pie, w = 4, h = 4)
print(pie)
share = intersect(P_gene, H_gene)
# writeLines(share, 'P-H.VDJ.share.txt')
rm(P_gene, H_gene, venn, pie)
cat('shared genes:', share[1:3], '...')
```
4. Compare shared genes
TRAV fisher.test:
```{r}
chain   = 'TRA'
genes   = grep(paste0(chain, 'V'), share, value = T)
PH      = TableVDJ(vdj, target = genes, names = c('P', 'H'), out.pref = 'PH.TRAV.share')
PH      = PH[order(-PH$P), c('Gene', 'P', 'H')]
PH      = cbind(PH, Fisher(PH$P, PH$H))
PH$Type = factor(ifelse(PH$Pval < .05, ifelse(PH$Log2Fc > 0, 'Up', 'Down'), 'Not'),
                 c('Up', 'Down', 'Not'))
# write.table(PH, paste0('P-H.', chain, 'V.fisher.txt'), sep = '\t', row.names = F)
rm(genes)
head(PH)
```
Volcano plot:
```{r}
suppressMessages(library(ggplot2))
p = ggplot(PH, aes(Log2Fc, -log10(Pval))) +
  geom_point(aes(color = Type)) +
  geom_hline(yintercept = -log10(.05), lty = 2, color = 'grey') +
  geom_vline(xintercept = 0, lty = 2, color = 'grey') +
  labs(x     = 'Fold Change (log2)',
       y     = expression(-log[10](Pvalue)),
       title = paste0(chain, 'V')) +
  scale_color_manual(values = c(if('Up'   %in% PH$Type) 'red',
                                if('Down' %in% PH$Type) 'blue', 'black')) +
  xlim(-1, 1) + ylim(0, 10) + theme_bw() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(hjust = .5))
suppressWarnings(ggsave(paste0('P-H.', chain, 'V.volcano.pdf'), p, w = 6, h = 5))
suppressWarnings(print(p))
```
Heatmap and fisher plot:
```{r}
suppressMessages(library(ggheatmap))
col_func  = colorRampPalette(rev(
  RColorBrewer::brewer.pal(n = 7, name = "RdYlBu") ))(100)
heatData  = setNames(PH[grep('proportion', names(PH))], c('P', 'H'))
rownames(heatData) = PH$Gene
# heatmap
p1 = ggheatmap(heatData,
               levels_rows        = rev(rownames(heatData)),
               levels_cols        = colnames(heatData),
               legendName         = 'Proportion',
               text_position_rows = 'left',
               border             = 'lightgrey',
               color              = col_func) +
  theme(legend.position           = 'left',
        legend.justification      = c(0, .05),
        legend.title              = element_text(angle = 90, hjust = .5),
        text                      = element_text(size = 13)) +
  guides(fill                     = guide_colorbar(title.position = 'left'))
rm(heatData)
# fisher
PH$Sig = ifelse(PH$Type == 'Not', 'Not', 'Sig')
p2 = ggplot(PH,  aes(Odds, Gene)) +
  geom_vline(xintercept = 1, lty = 2) +
  geom_errorbar(aes(xmin = ConfMin, xmax = ConfMax), width = .3, color = 'grey') +
  geom_point(aes(color = Sig), size = 4) +
  scale_color_manual(values = c(if('Sig' %in% PH$Sig) 'red', 'grey'),
                     labels = c(if('Sig' %in% PH$Sig) 'P < 0.05', 'P > 0.05')) +
  labs(y = '', x = '') +
  theme(axis.text.y          = element_blank(),
        axis.ticks.y         = element_blank(),
        axis.line.x          = element_line(),
        panel.grid           = element_blank(),
        panel.background     = element_blank(),
        text                 = element_text(size = 13),
        legend.key           = element_rect(fill = 'white'),
        legend.title         = element_blank(),
        legend.position      = 'right',
        legend.justification = c(1, 1))
p = p1 + p2
# ggsave(paste0('P-H.', chain, 'V.heatFisher.pdf'), p, width = 7, height = 9)
print(p)
```
Heatmap in samples:
```{r}
suppressMessages(library(ComplexHeatmap))
samples   = c(vdj@info$P, vdj@info$H)
PH_sig    = TableVDJ(vdj, target = PH$Gene[PH$Sig == 'Sig'], names = samples,
                     save = F)[seq(length(samples)+1)]
heatData  = t(apply(sweep(PH_sig[-1], 2, colSums(PH_sig[-1]), FUN = '/'), 1, scale))
col_func  = circlize::colorRamp2(c(-2,0,2), c('#116dac', 'white', '#b81e34'))
top_annot = HeatmapAnnotation(df  = data.frame(Group = sub('.$', '', samples)),
                              col = list(Group = c(P = '#b81e34', H = '#116dac')),
                              gap = unit(1, 'mm'),
                              show_legend          = F,
                              show_annotation_name = F )
p = Heatmap(heatData,
            col                   = col_func,
            rect_gp               = gpar(col = 'grey', lwd = 1),
            cluster_rows          = F,
            cluster_columns       = F,
            top_annotation        = top_annot,
            row_names_side        = 'right',
            column_names_rot      = 0,
            column_labels         = samples,
            row_labels            = PH_sig$Gene,
            column_names_side     = 'top',
            column_names_centered = T,
            heatmap_legend_param  = list(title = '', direction = 'horizontal'),
            cell_fun              = function(i, j, x, y, width, height, fill)
              grid.text(PH_sig[-1][j, i], x, y, gp = gpar(fontsize = 10)) )
# pdf(paste0('P-H.', chain, 'V.heatSamples.pdf'), width = 10, height = 8)
draw(p, heatmap_legend_side = 'bottom') # ; dev.off()
```
5. Filter vj-pair
Table VJ-pairs:
```{r}
P_vj = TableVJpair(vdj, names = vdj@info$P, out.pref = 'P')
H_vj = TableVJpair(vdj, names = vdj@info$H, out.pref = 'H')
head(H_vj)
```
Filter > 2 sample:
```{r}
P_vj = P_vj$VJ[P_vj$TotalName > 2]
H_vj = H_vj$VJ[H_vj$TotalName > 2]
# venn
venn   = VennDiagram::venn.diagram(list(P = P_vj, H = H_vj),
                                   col             = scales::hue_pal()(2),
                                   margin          = 0.05,
                                   fontfamily      = 'sans',
                                   cat.fontfamily  = 'sans',
                                   filename        = NULL,
                                   disable.logging = T)
grid::grid.draw(venn)
dev.off()
# pie
pie = Pie(P_vj, H_vj) + Pie(H_vj, P_vj, fill = c('white', '#00BFC4'))
# ggsave('P-H.VJpair.uniquePie.pdf', pie, w = 4, h = 4)
print(pie)
share = intersect(P_vj, H_vj)
# writeLines(share, 'P-H.VJpair.share.txt')
rm(P_vj, H_vj, venn, pie)
cat('shared VJ-pairs:', share[1:3], '...')
```
6. Compare shared vj-pairs
TRA V-J fisher.test:
```{r}
chain   = 'TRA'
genes   = grep(chain, share, value = T)
PH      = TableVJpair(vdj, target = genes, names = c('P', 'H'), out.pref = 'PH_TRA-VJ.share')
PH      = PH[order(-PH$P), c('VJ', 'P', 'H')]
PH      = cbind(PH, Fisher(PH$P, PH$H))
PH$Type = factor(ifelse(PH$Pval < .05, ifelse(PH$Log2Fc > 0, 'Up', 'Down'), 'Not'),
                 c('Up', 'Down', 'Not'))
# write.table(PH, paste0('P-H.', chain, '-VJ.fisher.txt'), sep = '\t', row.names = F)
head(PH)
```
Volcano plot:
```{r}
suppressMessages(library(ggplot2))
p = ggplot(PH, aes(Log2Fc, -log10(Pval))) +
  geom_point(aes(color = Type)) +
  geom_hline(yintercept = -log10(.05), lty = 2, color = 'grey') +
  geom_vline(xintercept = 0, lty = 2, color = 'grey') +
  labs(x     = 'Fold Change (log2)',
       y     = expression(-log[10](Pvalue)),
       title = paste(chain, 'V-J')) +
  scale_color_manual(values = c(if('Up'   %in% PH$Type) 'red',
                                if('Down' %in% PH$Type) 'blue', 'black')) +
  xlim(-4.5, 4.5) + ylim(0, 55) + theme_bw() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(hjust = .5))
suppressWarnings(ggsave(paste0('P-H.', chain, '-VJ.volcano.pdf'), p, w = 6, h = 5))
suppressWarnings(print(p))
```
Heatmap and fisher plot:
```{r}
suppressMessages(library(ggheatmap))
col_func  = colorRampPalette(rev(
  RColorBrewer::brewer.pal(n = 7, name = 'RdYlBu') ))(100)
heatData  = setNames(PH[grep('proportion', names(PH))], c('P', 'H'))
rownames(heatData) = PH$VJ
i         = seq(min(40, nrow(PH)))
heatData  = heatData[i, ]
p1 = ggheatmap(heatData, 
               levels_rows        = rev(rownames(heatData)),
               levels_cols        = colnames(heatData),
               legendName         = 'Proportion',
               text_position_rows = 'left',
               border             = 'lightgrey',
               color              = col_func) +
  theme(legend.position      = 'left',
        legend.justification = c(0, .05),
        legend.title         = element_text(angle = 90, hjust = .5),
        text                 = element_text(size = 13)) +
  guides(fill                = guide_colorbar(title.position = 'left'))
rm(heatData)
PH$Sig = ifelse(PH$Type == 'Not', 'Not', 'Sig')
p2 = ggplot(PH[i, ],  aes(Odds, VJ)) +
  geom_vline(xintercept = 1, lty = 2) +
  geom_errorbar(aes(xmin = ConfMin, xmax = ConfMax), width = .3, color = 'grey') +
  geom_point(aes(color = Sig), size = 4) +
  scale_color_manual(values = c(if('Sig' %in% PH$Sig[i]) 'red', 'grey'),
                     labels = c(if('Sig' %in% PH$Sig[i]) 'P < 0.05', 'P > 0.05')) +
  labs(y = '', x = '') +
  theme(axis.text.y          = element_blank(),
        axis.ticks.y         = element_blank(),
        axis.line.x          = element_line(),
        panel.grid           = element_blank(),
        panel.background     = element_blank(),
        text                 = element_text(size = 13),
        legend.key           = element_rect(fill = 'white'),
        legend.title         = element_blank(),
        legend.position      = 'right',
        legend.justification = c(1, 1))
p = p1 + p2
# ggsave(paste0('P-H.', chain, '-VJ.heatFisher.pdf'), p, width = 7, height = 9)
print(p)
```
Heatmap in samples:
```{r}
suppressMessages(library(ComplexHeatmap))
samples   = c(vdj@info$P, vdj@info$H)
PH_sig    = TableVJpair(vdj, target = PH$VJ[PH$Sig == 'Sig'], names = samples,
                        save = F)[seq(length(samples)+1)]
heatData  = t(apply(sweep(PH_sig[-1], 2, colSums(PH_sig[-1]), FUN = '/'), 1, scale))
col_func  = circlize::colorRamp2(c(-2,0,2), c('#116dac', 'white', '#b81e34'))
top_annot = HeatmapAnnotation(df  = data.frame(Group = sub('.$', '', samples)),
                              col = list(Group = c(P = '#b81e34', H = '#116dac')),
                              gap = unit(1, 'mm'),
                              show_legend          = F,
                              show_annotation_name = F )
p = Heatmap(heatData,
            col                   = col_func,
            rect_gp               = gpar(col = 'grey', lwd = 1),
            cluster_rows          = F,
            cluster_columns       = F,
            top_annotation        = top_annot,
            row_names_side        = 'right',
            column_names_rot      = 0,
            column_labels         = samples,
            row_labels            = PH_sig$VJ,
            column_names_side     = 'top',
            column_names_centered = T,
            heatmap_legend_param  = list(title = '', direction = 'horizontal'),
            cell_fun              = function(i, j, x, y, width, height, fill)
              grid.text(PH_sig[-1][j, i], x, y, gp = gpar(fontsize = 10)) )
# pdf(paste0('P-H.', chain, '-VJ.heatSamples.pdf'), width = 10, height = 8)
draw(p, heatmap_legend_side = 'bottom') # ; dev.off()
```
7. Filter vjAB
Table vjAB:
```{r}
P_ab = TableVJab(vdj, names = vdj@info$P, out.pref = 'P')
H_ab = TableVJab(vdj, names = vdj@info$H, out.pref = 'H')
head(H_ab)
```
Filter > 2 sample:
```{r}
P_ab = P_ab$VJab[P_ab$TotalName > 2]
H_ab = H_ab$VJab[H_ab$TotalName > 2]
# venn
venn   = VennDiagram::venn.diagram(list(P = P_ab, H = H_ab),
                                   col             = scales::hue_pal()(2),
                                   margin          = 0.05,
                                   fontfamily      = 'sans',
                                   cat.fontfamily  = 'sans',
                                   filename        = NULL,
                                   disable.logging = T)
grid::grid.draw(venn)
dev.off()
# pie
pie    = Pie(P_ab, H_ab) + Pie(H_ab, P_ab, fill = c('white', '#00BFC4'))
# ggsave('P-H.vjAB.uniquePie.pdf', pie, w = 4, h = 4)
print(pie)
P_uAB  = setdiff(P_ab, H_ab)
# writeLines(P_uAB, 'P-H.vjAB.unique.txt')
rm(P_ab, H_ab, venn, pie)
cat('differ VJab:', P_uAB[1:3], '...')
```
8. Compare virus
UpSet virus:
```{r}
Pu = TableVJab(vdj, target = P_uAB, names = c('P', vdj@info$V), out.pref = 'PV.unique')
head(Pu)
```

